Supermartingales and Termination Analysis of Statistical PCF

### 2021-10-01 LICS21/SPCF.tex
Use \ref instead of \Cref.
" \ref" -> "~\ref"
Various minors (formatting)

The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \citep{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,DBLP:conf/esop/MakOPW21}.
%\changed[lo]{following \cite{DBLP:conf/esop/MakOPW21} except the score function (\Cref{rem:score})}. 
Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
\end{align*}
The typing rules are standard (see \Cref{fig:typing rules}).
Terms are identified up to $\alpha$-equivalence, as usual. 
The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.

@@

%\paragraph{}
To define the CBV reduction relation, let \emph{evaluation contexts} be of the form:
\begin{align*}
  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2}
\end{align*}
then a term reduces if it is formed by substituting a redex in a context i.e.
\begin{align*}
  E[(\lambda x.M) V] & \to E[M[V/x]] \\
  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \\
  & \qquad \qquad \text{ where $z$ is not free in $M$}\\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1].
\end{align*}
We write $\to^\ast$ for the reflexive, transitive closure of $\to$.
Every closed term either is a value or reduces to another term.

>>>>

The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \citep{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,DBLP:conf/esop/MakOPW21}.
%\changed[lo]{following \cite{DBLP:conf/esop/MakOPW21} except the score function (\Cref{rem:score})}. 
Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
\end{align*}
The typing rules are standard (see Figure~\ref{fig:typing rules}).
Terms are identified up to $\alpha$-equivalence, as usual. 
The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.

@@

%\paragraph{}
To define the CBV reduction relation, let \emph{evaluation contexts} be of the form:
\[
  E ::= [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \mid \tY E \mid \tif{E<0}{M_1}{M_2}
\]
then a term reduces if it is formed by substituting a redex in a context i.e.
\begin{align*}
  E[(\lambda x.M) V] & \to E[M[V/x]] \\
  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z]\quad \text{ where $z$ is not free in $M$}\\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1].
\end{align*}
We write $\to^\ast$ for the reflexive, transitive closure of $\to$.
Every closed term either is a value or reduces to another term.

### 2021-10-01 LICS21/antitone.tex
Various minors (formatting)

\begin{example}[Unbiased random walk]
\label{ex:unbiased random walk}
For $M_2$, define the function $g_2 : \nnReal \to \nnReal$ by $g_2(x) = \ln(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone sparse ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
by
\(
f_2: {\Theta_2} \, n \mapsto 
g_2(n)
\)
and
$0 \mapsto 0$
for $n \in \mathbb N$.
% \begin{align*}
% {\Theta_2} \, n 
% &\mapsto 
% g_2(n)
% \\
% 0 &\mapsto 0.
% \end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to the Confluent Ranking Theorem~\ref{thm:confluent ranking}, as can the others in this section with similar issues. See Remark~\ref{rem:ex corrected using confluent semantics} for a fuller account.} each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}%\textstyle
& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) 
  =  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \ln\left(1 + \frac 1 {n(n+2)}\right) 
  >  \frac 1 {n(n+2) + 1} \\
  &>  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
\end{align*}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = 1 / {(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
\end{example}

@@

\begin{example}[More complex non-affine recursion]
\label{ex:non-affine continuous}
\begin{align*}
& (\tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample)\, \underline 0 \quad \hbox{where}\\
& X[e] := \tif{e \leq p - x^{-2}}{x}{f^2(x+1) \oplus_e f(x+1)}
\end{align*}
\end{example}

\begin{example}[Higher-order recursion] \label{ex:higher-order recursion}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \cite{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\qquad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\end{array}
\]
For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\, \underline n\, \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\, F\, \underline x\, \underline y$ is antitone rankable.
\end{example}

>>>>

\begin{example}[Unbiased random walk]
\label{ex:unbiased random walk}
For $M_2$, define the function $g_2 : \nnReal \to \nnReal$ by $g_2(x) = \ln(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone sparse ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
by
\(
f_2: {\Theta_2} \, n \mapsto 
g_2(n)
\)
and
$0 \mapsto 0$
for $n \in \mathbb N$.
% \begin{align*}
% {\Theta_2} \, n 
% &\mapsto 
% g_2(n)
% \\
% 0 &\mapsto 0.
% \end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to the Confluent Ranking Theorem~\ref{thm:confluent ranking}, as can the others in this section with similar issues. See Remark~\ref{rem:ex corrected using confluent semantics} for a fuller account.} each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}%\textstyle
g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 
  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) 
  =  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \ln\left(1 + \frac 1 {n(n+2)}\right) 
  >  \frac 1 {n(n+2) + 1} \\
  &>  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
\end{align*}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = 1 / {(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
\end{example}

@@

\begin{example}[More complex non-affine recursion]
\label{ex:non-affine continuous}
\iffalse
\begin{align*}
& (\tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample)\, \underline 0 \quad \hbox{where}\\
& X[e] := \tif{e \leq p - x^{-2}}{x}{f^2(x+1) \oplus_e f(x+1)}
\end{align*}
\fi
\[
(\tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample)\, \underline 0 \quad \hbox{where} \quad
X[e] := \tif{e \leq p - x^{-2}}{x}{f^2(x+1) \oplus_e f(x+1)}
\]
\end{example}

\begin{example}[Higher-order recursion] \label{ex:higher-order recursion}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \cite{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} \,.\, \tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\]
\iffalse
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\qquad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\end{array}
\]
\fi
For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\, \underline n\, \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\, F\, \underline x\, \underline y$ is antitone rankable.
\end{example}
